---
title: 'Lab 3: Panel Models'
subtitle: 'US Traffic Fatalities: 1980 - 2004'
output: 
  bookdown::pdf_document2: default
---

```{r load packages, echo=FALSE, message=FALSE}
library(tidyverse)
library(gridExtra)
library(tsibble)
library(forecast)
library(patchwork)
library(fable)
library(sandwich)
library(lmtest)
library(tseries)
library(gtrendsR)
library(quantmod)
library(vars)
library(GGally)
library(plm)
```


# U.S. traffic fatalities: 1980-2004

In this lab, we are asking you to answer the following **causal** question: 

> **"Do changes in traffic laws affect traffic fatalities?"**  

To answer this question, please complete the tasks specified below using the data provided in `data/driving.Rdata`. This data includes 25 years of data that cover changes in various state drunk driving, seat belt, and speed limit laws. 

Specifically, this data set contains data for the 48 continental U.S. states from 1980 through 2004. Various driving laws are indicated in the data set, such as the alcohol level at which drivers are considered legally intoxicated. There are also indicators for “per se” laws—where licenses can be revoked without a trial—and seat belt laws. A few economics and demographic variables are also included. The description of the each of the variables in the dataset is also provided in the dataset. 

```{r load data, echo = TRUE}
load(file="./data/driving.RData")

## please comment these calls in your work 
glimpse(data)
desc
```


# (30 points, total) Build and Describe the Data 

1. (5 points) Load the data and produce useful features. Specifically: 
    - Produce a new variable, called `speed_limit` that re-encodes the data that is in `sl55`, `sl65`, `sl70`, `sl75`, and `slnone`; 
    - Produce a new variable, called `year_of_observation` that re-encodes the data that is in `d80`, `d81`, ... , `d04`. 
    - Produce a new variable for each of the other variables that are one-hot encoded (i.e. `bac*` variable series). 
    - Rename these variables to sensible names that are legible to a reader of your analysis. For example, the dependent variable as provided is called, `totfatrte`. Pick something more sensible, like, `total_fatalities_rate`. There are few enough of these variables to change, that you should change them for all the variables in the data. (You will thank yourself later.)

```{r features creation, warning=FALSE, echo=FALSE, message=FALSE}
data <- data %>% mutate(
  speed_limit = sl55 * 55 + sl65 * 65 + sl70 * 70 + sl75 * 75 + slnone * 0,
  speed_limit_bin = case_when((speed_limit <55 ) ~ "No speed limit",
                              (speed_limit >= 55) & (speed_limit < 65) ~ "Between 55 and 65",
                              (speed_limit >= 65) & (speed_limit < 70) ~ "Between 65 and 70",
                              (speed_limit >= 70) & (speed_limit < 75) ~ "Between 70 and 75",
                              (speed_limit >= 75) ~ "Greater than 75"),
  year_of_observation = year * (d80 + d81 + d82 + d83 + d84 + d85 + d86 + d87 +
                                d88 + d89 + d90 + d91 + d92 + d93 + d94 + d95 +
                                d96 + d97 + d98 + d99 + d00 + d01 + d02 + d03 + d04),
  year_of_observation = as.character(year_of_observation),
  bld_alc_lmt = bac10 * 0.1 + bac08 * 0.08
)

data <- rename(data, speed_lim_55 = sl55,
                     speed_lim_65 = sl65,
                     speed_lim_70 = sl70,
                     speed_lim_75 = sl75,
                     no_speed_lim = slnone,
                     seatbelt_law = seatbelt,
                     min_drink_age = minage,
                     zero_tol_law = zerotol,
                     grad_driver_license_law = gdl,
                     bld_alc_lim10 = bac10,
                     bld_alc_lim08 = bac08,
                     per_se_law = perse,
                     tot_traffic_fata = totfat,
                     tot_night_fata = nghtfat,
                     tot_wkend_fata = wkndfat,
                     night_fata_per_100milmile = nghtfatpvm,
                     wkend_fata_per_100milmile = wkndfatpvm,
                     state_pop = statepop,
                     tot_fata_per_100th_pop = totfatrte,
                     night_fata_per_100th_pop = nghtfatrte,
                     wkend_fata_per_100th_pop = wkndfatrte,
                     veh_mile_trav = vehicmiles,
                     unemp_rate = unem,
                     age1424_pop_percent = perc14_24,
                     speed_lim_grter70 = sl70plus,
                     prim_seatbelt_law = sbprim,
                     second_seatbelt_law = sbsecon) 
```

2. (5 points) Provide a description of the basic structure of the dataset. What is this data? How, where, and when is it collected? Is the data generated through a survey or some other method? Is the data that is presented a sample from the population, or is it a *census* that represents the entire population? Minimally, this should include:
    - How is the our dependent variable of interest `total_fatalities_rate` defined? 
    
The dataset contains the total car fatalities rate from 1980 to 2004 for the 48 states in the US. The data is structured under a panel format. It is believed that the data was collected by National Highway Traffice Safety Administration (NHTSA). According to NHTSA, the traffic fatilities data is obtained from States' existing documents: police accident reports, state vehicle registration files and state driver licensing files. When dividing the total traffic fatalities by the state population, we obtained the result similar to total traffic fatality rate (fatalities per 100,000 of population), so it's likely the data represent the entire population within the 48 states. 

As mentioned above, the `total_fatilities_rate` is the total traffic fatalities per 100,000 of population. A `total_facilities_rate` of 24 means for every 100,000 residents within the state, there are 24 fatalities due to traffic. 

Besides fatilities data, the dataset also consists of traffic law data (speed limit, blood alcohol limit, per se law, seatbelt law, zero tolerance law), and other data that may correlate with traffic fatilities (minimum drinking age, vehicle travel miles, percentage of population with age from 14 to 24 and unemployment rate). 

3. (20 points) Conduct a very thorough EDA, which should include both graphical and tabular techniques, on the dataset, including both the dependent variable `total_fatalities_rate` and the potential explanatory variables. Minimally, this should include: 
    - How is the our dependent variable of interest `total_fatalities_rate` defined? 
    - What is the average of `total_fatalities_rate` in each of the years in the time period covered in this dataset? 

As with every EDA this semester, the goal of this EDA is not to document your own process of discovery -- save that for an exploration notebook -- but instead it is to bring a reader that is new to the data to a full understanding of the important features of your data as quickly as possible. In order to do this, your EDA should include a detailed, orderly narrative description of what you want your reader to know. Do not include any output -- tables, plots, or statistics -- that you do not intend to write about.

```{r EDA, message=FALSE, echo=FALSE, warning=FALSE}
data %>% ggplot(aes(x = year, y = tot_fata_per_100th_pop)) + geom_line() + ggtitle("Total Trafic Fatalities Over Years")
data %>% ggplot(aes(x = year, y = tot_fata_per_100th_pop)) + geom_line() + facet_wrap(~state) + ggtitle("Total Trafic Fatilities Over Years by States")
data %>% ggplot() + geom_boxplot(aes(y = tot_fata_per_100th_pop, x = factor(state))) + ggtitle("Total Trafic Fatalities by States")
data %>% ggplot() + geom_boxplot(aes(y = tot_fata_per_100th_pop, x = factor(year))) + ggtitle("Total Trafic Fatalities by Year")
data %>% ggplot() + geom_boxplot(aes(y = tot_fata_per_100th_pop, x = factor(speed_limit_bin))) + ggtitle("Total Trafic Fatalitites by Speed Limit Category")
data2 <- data %>% dplyr::select(tot_fata_per_100th_pop, night_fata_per_100th_pop, wkend_fata_per_100th_pop,
                         unemp_rate, age1424_pop_percent, veh_mile_trav)
ggpairs(data2, columns = 1:6)
data %>% ggplot() + geom_point(aes(y = tot_fata_per_100th_pop, x = night_fata_per_100th_pop)) + geom_smooth(aes(y = tot_fata_per_100th_pop, x = night_fata_per_100th_pop)) + facet_wrap(~state) + ggtitle("Total Fatalities vs Night Fatalities by State")
data %>% ggplot() + geom_point(aes(y = tot_fata_per_100th_pop, x = wkend_fata_per_100th_pop)) + geom_smooth(aes(y = tot_fata_per_100th_pop, x = wkend_fata_per_100th_pop)) + facet_wrap(~state) + ggtitle("Total Fatalities vs Weekend Fatalities by State")
data %>% ggplot() + geom_point(aes(y = tot_fata_per_100th_pop, x = unemp_rate)) + geom_smooth(aes(y = tot_fata_per_100th_pop, x = unemp_rate)) + facet_wrap(~state) + ggtitle("Total Fatalities vs Unemployment by State")
data %>% ggplot() + geom_point(aes(y = tot_fata_per_100th_pop, x = age1424_pop_percent)) + geom_smooth(aes(y = tot_fata_per_100th_pop, x = age1424_pop_percent)) + facet_wrap(~state) + ggtitle("Total Fatalities vs Percent of People Younger than 24 by State")
data %>% ggplot() + geom_point(aes(y = tot_fata_per_100th_pop, x = veh_mile_trav)) + geom_smooth(aes(y = tot_fata_per_100th_pop, x = veh_mile_trav)) + facet_wrap(~state) + ggtitle("Total Fatalities vs Miles Traveling")
data %>% ggplot() + geom_boxplot(aes(y = tot_fata_per_100th_pop, x = factor(bld_alc_lmt))) + ggtitle("Total Fatilities by Blood Alcohol Limit")

data %>% ggplot() + geom_boxplot(aes(y = tot_fata_per_100th_pop, x = factor(per_se_law))) + ggtitle("Total Fatalities based on Per Se Law")
data %>% ggplot() + geom_boxplot(aes(y = tot_fata_per_100th_pop, x = factor(seatbelt_law))) + ggtitle("Total Fatalities based on Seatbelt Law")
data %>% ggplot() + geom_boxplot(aes(y = tot_fata_per_100th_pop, x = factor(zero_tol_law))) + ggtitle("Total Fatalities based on Zero Tolerance Law")

```

The above plots show the following: 
 - There is a high correlation between total traffic fatalities fatalities during night and weekend. This correlation appears to be similar among all states. 
 - Traffic fatalities are lower when there is law enforced (applicable to all laws included in the dataset, per se law, seat belt law and zero tolerance law)
 - There is a reduction in traffic fatalities from 1980 to 2004, although it could be due to the enforcement of trafic law
 - The downward trend of traffic fatalities seems to follow for all states
 - There do not seem to have high correlation between traffic fatalities with miles traveling, percent of people younger than 24 and unemployment rate for all states. There are exception but it maybe spurious correlation

# (15 points) Preliminary Model

Estimate a linear regression model of *totfatrte* on a set of dummy variables for the years 1981 through 2004 and interpret what you observe. In this section, you should address the following tasks: 

- Why is fitting a linear model a sensible starting place? 
- What does this model explain, and what do you find in this model? 
- Did driving become safer over this period? Please provide a detailed explanation.
- What, if any, are the limitation of this model. In answering this, please consider **at least**: 
    - Are the parameter estimates reliable, unbiased estimates of the truth? Or, are they biased due to the way that the data is structured?
    - Are the uncertainty estimate reliable, unbiased estimates of sampling based variability? Or, are they biased due to the way that the data is structured? 

```{r Preliminary Model, echo = FALSE, warning = FALSE, message = FALSE}
pdata <- pdata.frame(data, index = c("year", "state")) 
pre_model <- plm(tot_fata_per_100th_pop ~ d80 + d81 + d82 + d83 + d84 + d85 + d86 + d87 + d88 + d89 + d90 +
                   d91 + d92 + d93 + d94 + d95 + d96 + d97 + d98 + d99 + d00 + d01 + d02 + d03 + d04, data = pdata, 
                 index = c("year", "state"), 
                 effect = "individual", model = "pooling")
summary(pre_model)


plot(a[1:nrow(a),2], a[1:nrow(a),1], xlab = "Fitted Value", ylab = "Residuals") 


```
The EDA shows a series of correlation between the response variable (total traffic fatality rate) and explanatory variables. As a result, a linear model is an appropriate approach to forecast the total traffic fatality rate. 

The model uses `d04` (2004 indicator) as the base. Based on the output of the model, each and every year prior to 2004 has a higher fatality rate. From 1980 to 1990, the p-values of the coefficient estimates show that the difference in fatalities in these years is statistically significant. However from 1991, the coefficient estimates are no longer statistically significant. This suggests that the fatality rate does not just change overtime and there are other omitted variables that impact the fatality rates. 

Because it's likely that there are omitted variables within the model, the parameter estimates are not reliable and likely overestimate/underestimate the impact of time on fatalities rate. One omitted variable that we suspect is the enforcement of traffic law which happened in the later years (1980 as the reference). The enforced traffic law would likely reduce the fatalities rate (as shown in the EDA section), which would imply that the impact from years on traffic fatalities is less than what the above model shows. 

The residual versus fitted values plot shows that there is a relationship between the model residuals and the fitted values. This suggests that the model uncertainty estimates are not reliable and maybe biased. 


# (15 points) Expanded Model 

Expand the **Preliminary Model** by adding variables related to the following concepts: 

- Blood alcohol levels 
- Per se laws
- Primary seat belt laws (Note that if a law was enacted sometime within a year the fraction of the year is recorded in place of the zero-one indicator.)
- Secondary seat belt laws 
- Speed limits faster than 70 
- Graduated drivers licenses 
- Percent of the population between 14 and 24 years old
- Unemployment rate
- Vehicle miles driven per capita. 

If it is appropriate, include transformations of these variables. Please carefully explain carefully your rationale, which should be based on your EDA, behind any transformation you made. If no transformation is made, explain why transformation is not needed. 

- How are the blood alcohol variables defined? Interpret the coefficients that you estimate for this concept. 
- Do *per se laws* have a negative effect on the fatality rate? 
- Does having a primary seat belt law? 

```{r Expand model, echo = FALSE, message = FALSE, warning = FALSE}

```

# (15 points) State-Level Fixed Effects 

Re-estimate the **Expanded Model** using fixed effects at the state level. 

- What do you estimate for coefficients on the blood alcohol variables? How do the coefficients on the blood alcohol variables change, if at all? 
- What do you estimate for coefficients on per se laws? How do the coefficients on per se laws change, if at all? 
- What do you estimate for coefficients on primary seat-belt laws? How do the coefficients on primary seatbelt laws change, if at all? 

Which set of estimates do you think is more reliable? Why do you think this? 

- What assumptions are needed in each of these models?  
- Are these assumptions reasonable in the current context?

# (10 points) Consider a Random Effects Model 

Instead of estimating a fixed effects model, should you have estimated a random effects model?

- Please state the assumptions of a random effects model, and evaluate whether these assumptions are met in the data. 
- If the assumptions are, in fact, met in the data, then estimate a random effects model and interpret the coefficients of this model. Comment on how, if at all, the estimates from this model have changed compared to the fixed effects model. 
- If the assumptions are **not** met, then do not estimate the data. But, also comment on what the consequences would be if you were to *inappropriately* estimate a random effects model. Would your coefficient estimates be biased or not? Would your standard error estimates be biased or not? Or, would there be some other problem that might arise?

# (10 points) Model Forecasts 

The COVID-19 pandemic dramatically changed patterns of driving. Find data (and include this data in your analysis, here) that includes some measure of vehicle miles driven in the US. Your data should at least cover the period from January 2018 to as current as possible. With this data, produce the following statements: 

- Comparing monthly miles driven in 2018 to the same months during the pandemic: 
  - What month demonstrated the largest decrease in driving? How much, in percentage terms, lower was this driving? 
  - What month demonstrated the largest increase in driving? How much, in percentage terms, higher was this driving? 
  
Now, use these changes in driving to make forecasts from your models. 

- Suppose that the number of miles driven per capita, increased by as much as the COVID boom. Using the FE estimates, what would the consequences be on the number of traffic fatalities? Please interpret the estimate.
- Suppose that the number of miles driven per capita, decreased by as much as the COVID bust. Using the FE estimates, what would the consequences be on the number of traffic fatalities? Please interpret the estimate.

# (5 points) Evaluate Error 

If there were serial correlation or heteroskedasticity in the idiosyncratic errors of the model, what would be the consequences on the estimators and their standard errors? Is there any serial correlation or heteroskedasticity? 